Agda Meets Accelerate

### 2012-11-30 paper/main.tex

This paper is a first investigation into this topic. It reports on a
partial port of Accelerate to a new, dependently-typed host language,
Agda \cite{Norell2008,BoveDybjerNorell2009}. Agda is particularly
suited to this port because of its foreign function interface to
Haskell, which enables it to directly invoke the functionality of
Accelerate.

Our investigation has the following structure. After recalling some
background on Agda and Accelerate in Section~\ref{sec:background} and
describing related work in Section~\ref{sec:related-work}, 
Section~\ref{sec:dependent-types} discusses potential uses of
dependent types in an array-oriented data parallel language and how they were realized in our
implementation. Section~\ref{sec:limitations} considers conceptual 
problems and limitations that we ran into when constructing the Agda
frontend for Accelerate. Section~\ref{sec:implementation} explains
some technical details of the implementation and discusses some
example code.

@@

The access operation takes a vector of length \verb+n+, an index
\verb+m+, and a proof of \verb+m < n+ (a derivation tree) to produce
an element of the vector.  
\begin{verbatim}
get : {A : Set} {n : Nat} -> Vec A n -> (m : Nat) -> m < n -> A
get []        _       ()      -- impossible case
get (x :: xs) zero    z<s     = x
get (x :: xs) (suc m) (s<s p) = get xs m p
\end{verbatim}
This code cannot fail at run time because a caller has to
construct the proof tree for \verb+m < n+ before invoking
\verb+get+. (In Agda, arguments in curly braces are \emph{implicit
  arguments} that will be inferred if omitted in an application.)
Thus, an ``index out of bounds'' error cannot happen.

@@

Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a
data-parallel array language embedded into Haskell, which targets
GPUs. It is a \emph{generative library}, as its data-parallel array
operations are not executed directly, but instead construct abstract
syntax trees (AST) representing an entire data-parallel
subcomputation. These \emph{computation representations} are executed
using a \verb+run+ operation that accepts such a representation (of type
\verb+Acc a+), compiles it to GPU kernels, uploads it to the device,
executes it, and retrieves the results.\footnote{To distinguish
  Haskell code from Agda code, we display Haskell code in blue and with
  a vertical bar on the left side.}
%
\begin{hcode}
CUDA.run :: Arrays a => Acc a -> a
\end{hcode}
%
The type class constraint \verb+Arrays a+ restricts the result type to a single array or a tuple of arrays.

@@

Accelerate supports a wide range of numeric types, characterized by
type class \texttt{Elt}, as base types for
array computations. Almost all of these types have no suitable
counterpart in Agda, which only supplies  computationally expensive
encodings for natural and rational numbers. For that reason, our embedding keeps the Haskell types 
abstract in Agda. To specify the types of functions that are polymorphic in such a Haskell type or depend on it
in some way, we have reified the possible element types in an Agda type
\texttt{Elt}.
\begin{verbatim}
data Elt : Set where
  Bool   : Elt
  Int    : Elt
  Float  : Elt
  Double : Elt
  Pair   : Elt -> Elt -> Elt
  -- and so on
\end{verbatim}
Corresponding to Haskell type classes that are used in Accelerate, our
embedding supplies predicates that characterize subsets. For example,
the set of numeric types is defined by a predicate
\texttt{Numeric}.\footnote{$\top$ is a one-element type,
  whereas $\bot$ is a type without elements. These types customarily
  represent truth and falsity.}
\begin{alltt}
Numeric : Elt -> Set
Numeric Int = \(\top\)
Numeric Float = \(\top\)
Numeric Double = \(\top\)
Numeric _ = \(\bot\)
\end{alltt}
The embedding declares further subsets all in the same style.

@@

However, an alternative encoding of arrays can be
used which is compatible with filtering of elements. The idea of this
encoding is to keep all elements but mark those which are no longer
present because they have been filtered out.
There are several ways of implementing this idea. The simplest
approach is to pair up each element with a boolean flag that indicates
its presence.\footnote{Using an array of \texttt{Maybe}-typed elements is not an option
  because Accelerate does not support \texttt{Maybe} types as array elements.}
\begin{verbatim}
FVector : Nat -> Elt -> Set
FVector n E = Vector n (Pair Bool E)
\end{verbatim}

In this encoding, filtering is quite simple because the length of
the \texttt{FVector} does not change. Furthermore, filtering could be
extended to multi-dimensional arrays, although it is not clear if an
array that contains non-existing elements or holes is a sensible notion.
\begin{verbatim}
filterF : forall {n : Nat}{E : Elt}
        -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
filterF {n}{E} pred vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool E)
  g bx = pair ((fst bx) && p (snd bx)) x
\end{verbatim}
The mapping operation that applies a function to each element of an
array becomes more complicated because it either has to 
materialize a dummy result for each absent element in the argument
vector or apply the function to absent elements, too.
\begin{verbatim}
mapF : forall {n : Nat}{E F : Elt}
     -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
mapF {n}{E}{F} defaultF f vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool F)
  g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
                     else (pair (fst bx) defaultF)
\end{verbatim}
On the positive side, some operations can get rid of the absent
elements. In particular, a fold operation which reduces a filtered
vector with a monoid returns a single value. In Accelerate, such a value has type
\texttt{Scalar}, which is a synonym for an array of dimension $0$.
\begin{verbatim}
foldF : forall {n : Nat}{E : Elt}
      -> (Exp E -> Exp E -> Exp E) -> Exp E
      -> FVector n E -> Scalar E
foldF f e vec =
  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
\end{verbatim}
Other operations like \texttt{fold1} and the scan operations present
in Accelerate can also be lifted to this representation, but they
retain a notion of absent elements and do not allow to revert to a
non-filtered representation.

@@

As an example for a function that requires more work on either side,
consider the \texttt{fold} operation. 
\begin{code}
fold : forall {E}{sh}{n}
     -> (Exp E -> Exp E -> Exp E)
     -> Exp E
     -> Array (sh :< n >) E
     -> Array sh E
fold {E}{sh}{n} f (Ex e) (Ar a) =
  Ar (foldHs (EltDict E) (toHsInt (size sh)) (toHsInt n)
             (unwrap2 f) e a)
\end{code}
As values of type \texttt{Exp} also need a wrapper type in Agda (it is
not possible to import type constructors via the FFI), there is some
unwrapping going on for the \texttt{e} and \texttt{f}
arguments. The implementation of \texttt{fold} just calls the
\texttt{foldHs} function and encodes the information about the shape
in two integer arguments. Here, \texttt{size sh} is the size of the
result and \texttt{n} is the size of the dimension that is folded. As
these values are initially available as Agda natural numbers, they need to
be converted to Haskell numbers using the function \texttt{toHsInt}.

>>>>

This paper is a first investigation into this topic. It reports on a
partial port of Accelerate to a new, dependently-typed host language,
Agda \cite{Norell2008,BoveDybjerNorell2009}. Agda is particularly
suited to this port because of its foreign function interface to
Haskell, which enables it to directly invoke the functionality of
Accelerate. The main contributions of this paper are the following:
%
\begin{itemize}
\item We identify and discuss the challenges of combining generative embedded languages with dependent typing (Section~\ref{sec:dependent-types}).
\item We propose predicated arrays to overcome some of these challenges (Section~\ref{sec:limitations}).
\item We outline an implementation of the main parts of Accelerate in Agda using the Agda-Haskell FFI for code execution (Section~\ref{sec:implementation}).
\end{itemize}
%
Overall, our investigation has the following structure. After recalling some
background on Agda and Accelerate in Section~\ref{sec:background} and
describing related work in Section~\ref{sec:related-work}, 
Section~\ref{sec:dependent-types} discusses potential uses of
dependent types in an array-oriented data parallel language and how they were realized in our
implementation. Section~\ref{sec:limitations} considers conceptual 
problems and limitations that we ran into when constructing the Agda
frontend for Accelerate. Section~\ref{sec:implementation} explains
some technical details of the implementation and discusses some
example code.

@@

The access operation takes a vector of length \verb+n+, an index
\verb+m+, and a proof of \verb+m < n+ (a derivation tree) to produce
an element of the vector.  
\begin{verbatim}
get : {A : Set} {n : Nat} -> Vec A n -> (m : Nat) -> m < n -> A
get []        _       ()      -- impossible case
get (x :: xs) zero    z<s     = x
get (x :: xs) (suc m) (s<s p) = get xs m p
\end{verbatim}
This code cannot fail at run time because a caller has to
construct the proof tree for \verb+m < n+ before invoking
\verb+get+. Thus, an ``index out of bounds'' error cannot happen. 
(In Agda, arguments in curly braces are \emph{implicit
  arguments} that will be inferred if omitted in an application.)

@@

Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a
data-parallel array language embedded into Haskell, which targets
GPUs. It is a \emph{generative library}, as its data-parallel array
operations are not executed directly, but instead construct abstract
syntax trees (AST) representing an entire data-parallel
subcomputation. These \emph{computation representations} are executed
using a \verb+run+ operation that accepts such a representation (of type
\verb+Acc a+), compiles it to GPU kernels, uploads it to a device,
executes it, and retrieves the results.\footnote{To distinguish
  Haskell code from Agda code, we display Haskell code in blue and with
  a vertical bar on the left side.}
%
\begin{hcode}
CUDA.run :: Arrays a => Acc a -> a
\end{hcode}
%
The type class constraint \verb+Arrays a+ restricts the result type to a single array or a tuple of arrays.

@@

Accelerate supports a wide range of numeric types, characterized by
the type class \texttt{Elt}, as base types for
array computations. Almost all of these types lack a suitable
counterpart in Agda, which only supplies  computationally expensive
encodings for natural and rational numbers. For that reason, our embedding keeps the Haskell types 
abstract in Agda. To specify the types of functions that are polymorphic in such a Haskell type or depend on it
in some way, we have reified the possible element types as an Agda type
\texttt{Elt}:
\begin{verbatim}
data Elt : Set where
  Bool   : Elt
  Int    : Elt
  Float  : Elt
  Double : Elt
  Pair   : Elt -> Elt -> Elt
  -- and so on
\end{verbatim}
Corresponding to Haskell type classes that are used in Accelerate, our
embedding supplies predicates that characterize subsets. For example,
the set of numeric types is defined by a predicate
\texttt{Numeric}:\footnote{$\top$ is a one-element type,
  whereas $\bot$ is a type without elements. These types customarily
  represent truth and falsity.}
\begin{alltt}
Numeric : Elt -> Set
Numeric Int = \(\top\)
Numeric Float = \(\top\)
Numeric Double = \(\top\)
Numeric _ = \(\bot\)
\end{alltt}
The embedding declares further subsets all in the same style.

@@

However, an alternative encoding of arrays can be
used which is compatible with filtering of elements. The idea of this
encoding is to keep all elements but mark those which are no longer
present because they have been filtered out.
There are several ways of implementing this idea. The simplest
approach is to pair up each element with a boolean flag that indicates
its presence, which we call \emph{predicated arrays}:\footnote{Using an array of \texttt{Maybe}-typed elements is not an option
  because Accelerate does not support \texttt{Maybe} types as array elements.}
\begin{verbatim}
FVector : Nat -> Elt -> Set
FVector n E = Vector n (Pair Bool E)
\end{verbatim}

In this encoding, filtering is quite simple because the length of
the \texttt{FVector} does not change. Furthermore, filtering could be
extended to multi-dimensional arrays, although it is not clear if an
array that contains non-existing elements or holes is a sensible notion.
\begin{verbatim}
filterF : forall {n : Nat}{E : Elt}
        -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
filterF {n}{E} pred vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool E)
  g bx = pair ((fst bx) && p (snd bx)) x
\end{verbatim}
The mapping operation that applies a function to each element of an
array becomes more complicated because it either has to 
materialize a dummy result for each absent element in the argument
vector or apply the function to absent elements, too. This makes \texttt{filter} reminiscent of the \texttt{where} statement of the SIMD language C$^*$~\cite{rose-etal:c-star}.
\begin{verbatim}
mapF : forall {n : Nat}{E F : Elt}
     -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
mapF {n}{E}{F} defaultF f vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool F)
  g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
                     else (pair (fst bx) defaultF)
\end{verbatim}
On the positive side, some operations can get rid of the absent
elements. In particular, a fold operation which reduces a filtered
vector with a monoid returns a single value. In Accelerate, such a value has type
\texttt{Scalar}, which is a synonym for an array of dimension $0$.
\begin{verbatim}
foldF : forall {n : Nat}{E : Elt}
      -> (Exp E -> Exp E -> Exp E) -> Exp E
      -> FVector n E -> Scalar E
foldF f e vec =
  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
\end{verbatim}
Other operations like \texttt{fold1} and the scan operations present
in Accelerate can also be lifted to this representation, but they
retain a notion of absent elements and do not allow to revert to a
non-filtered representation.

@@

As an example for a function that requires more work on both sides,
consider the \texttt{fold} operation. 
\begin{code}
fold : forall {E}{sh}{n}
     -> (Exp E -> Exp E -> Exp E)
     -> Exp E
     -> Array (sh :< n >) E
     -> Array sh E
fold {E}{sh}{n} f (Ex e) (Ar a) =
  Ar (foldHs (EltDict E) (toHsInt (size sh)) (toHsInt n)
             (unwrap2 f) e a)
\end{code}
As values of type \texttt{Exp} also need a wrapper type in Agda (it is
not possible to import type constructors via the FFI), there is some
unwrapping going on for the \texttt{e} and \texttt{f}
arguments. The implementation of \texttt{fold} just calls the
\texttt{foldHs} function and encodes the information about the shape
in two integer arguments. Here, \texttt{size sh} is the size of the
result and \texttt{n} is the size of the dimension that is folded. As
these values are initially available as Agda natural numbers, they need to
be converted to Haskell numbers using the function \texttt{toHsInt}.
